---
title: "2. Data cubes from Sentinel-2 data in the cloud"
author: "Marius Appel"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Data cubes from Sentinel-2 data in the cloud}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
ev = unname(Sys.info()["user"]) == "marius"
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = ev
)
```


Many cloud computing providers contain large catalogs of satellite imagery as from the Sentinel-2 satellites. To avoid downloading hundreds of images (and gigabytes) from portals, one can directly run a machine on a cloud platform and run the analysis without any downloads before. Fortunately, this has recently been made much easier and more efficient due to the development and popularity of the [cloud-optimized GeoTIFF](https://www.cogeo.org) image format and the [SpatioTemporal Asset Catalog (STAC)](https://stacspec.org).

This vignette will not explain any details the specifications but demonstrate how they can be used in combination with gdalcubes and the [`rstac`](https://cran.r-project.org/package=rstac) package. We will use the freely available [Sentinel-2 COG catalog](https://registry.opendata.aws/sentinel-2-l2a-cogs) on Amazon Web Services (AWS) and the corresponding STAC-API endpoint at https://earth-search.aws.element84.com/v0/collections/sentinel-s2-l2a.

Notice that this vignette in principle runs anywhere but computations take much longer when not working on an AWS machine in the region of the data catalog (in this case us-west-2).



# Finding images with `rstac`

As an example, we are interested in cloud-free images of New York City in June, 2021. We can use the 
NYC administrative polygons that are shipped as example data with the package and use the (`sf`)[https://cran.r-project.org/package=sf] package for reading and calculating the bounding box.

```{r}
library(sf)
nyc_shape = read_sf(system.file("nycd.gpkg",package = "gdalcubes"))

bbox = st_bbox(nyc_shape) 
bbox
```

To ask the STACK endpoints for available images intersecting with our area of interest, we need to provide the bounding box in WGS84 (latitude / longitude) coordinates, which we can calculate with: 


```{r}
nyc_shape |>
  st_transform("EPSG:4326") |>
  st_bbox() -> bbox_wgs84
bbox_wgs84
```

Next, we load the `rstac` package, specify the STAC-API endpoint URL and query all available images for our area and time of interest.

```{r}
library(rstac)
s = stac("https://earth-search.aws.element84.com/v0")
  items = s |>
    stac_search(collections = "sentinel-s2-l2a-cogs",
                bbox = c(bbox_wgs84["xmin"],bbox_wgs84["ymin"],
                         bbox_wgs84["xmax"],bbox_wgs84["ymax"]), 
                datetime = "2021-06-01/2021-06-30") |>
    post_request() |> items_fetch(progress = FALSE)
  length(items$features)
```
As a result, we get a list with metadata and URLs pointing to 48 images. This list is an index-like object pointing to original images on a AWS S3 bucket and hence somewhat similar to gdalcubes image collections.


# Converting STAC items to image collections


We can convert the STAC response to a gdalcubes image collections using `stac_image_collection()`. 
This function expects a STAC feature list as input and optionally can apply some filters on metadata and bands. Notice that this operation is much faster than the creation of image collections from local files, because all metadata are already available and no image file must be opened. Below, we create a collection for images with less than 10% cloud cover. 

```{r}
library(gdalcubes)
s2_collection = stac_image_collection(items$features,  property_filter = function(x) {x[["eo:cloud_cover"]] < 10})
s2_collection
```

The collection contains 16 images and all spectral bands plus some RGB preview images. However, the [scene classification layer (SCL)](https://sentinels.copernicus.eu/web/sentinel/technical-guides/sentinel-2-msi/level-2a/algorithm) containing pixel-wise information whether a pixel shows a cloud, cloud-shadow, water, or something else, is missing. Although the SCL images are available in the returned STAC list, the response does not include all of the metadata to let gdalcubes recognize it as an image. However, it is possible to explicitly list the names of all bands to be included in the collection by adding the `asset_names` argument:

```{r, warning=FALSE}
assets = c("B01","B02","B03","B04","B05","B06", "B07","B08","B8A","B09","B11","SCL")
s2_collection = stac_image_collection(items$features, asset_names = assets, property_filter =
                                      function(x) {x[["eo:cloud_cover"]] < 10})
s2_collection
```



# Creating and processing data cubes

Having an image collection object, we can now use gdalcubes in the same way as we do with local files. Particularly, we define a data cube view and maybe some additional data cube operations. In the following example, we create a coarse resolution (100m) simple median-composite RGB image from a daily data cube.

```{r, fig.align="center", fig.dim=c(4.5,4.5)}
gdalcubes_options(parallel = 8)
v = cube_view(srs="EPSG:32618", dx=100, dy=100, dt="P1D", 
                           aggregation="median", resampling = "average",
                           extent=list(t0 = "2021-06-01", t1 = "2021-06-30",
                                       left=bbox["xmin"], right=bbox["xmax"],
                                       top=bbox["ymax"], bottom=bbox["ymin"]))
v

raster_cube(s2_collection, v) |>
  select_bands(c("B02","B03","B04")) |>
  reduce_time(c("median(B02)", "median(B03)", "median(B04)")) |>
  plot(rgb = 3:1, zlim = c(0,2500))
```
Of course, we can "zoom in" to Lower Manhattan simply by changing the _data cube view_.

```{r, fig.align="center", fig.dim=c(4.5,4.5)}
v = cube_view(srs="EPSG:32618", dx=10, dy=10, dt="P1D", 
                           aggregation="median", resampling = "average",
                           extent=list(t0 = "2021-06-01", t1 = "2021-06-30",
                                       left=582182, right=587019,
                                       top=4508997, bottom=4505883))
v
raster_cube(s2_collection, v) |>
  select_bands(c("B02","B03","B04")) |>
  reduce_time(c("median(B02)", "median(B03)", "median(B04)")) |>
  plot(rgb = 3:1, zlim = c(0,2500))
```

For more complex examples, you can find a [tutorial on YouTube](https://youtu.be/Xlg__2PeTXM?t=3693) and corresponding materials on [GitHub](https://github.com/appelmar/ogh2021) how to use gdalcubes in the cloud, presented at the virtual OpenGeoHub Summer School 2021.

# Summary

Thanks to STAC, cloud-optimized GeoTIFFs, and GDAL being capable of reading imagery from cloud storage directly, moving to cloud-based analysis workflows without downloading any imagery from portals become a lot easier and more efficient. Whether or not this is the right approach depends a lot on particular applications. In some cases (e.g. when having access to institutional HPC resources while applying very complex models on small areas of interest), it might still be appropriate to download the data once. The `gdalcubes` package still can help as an interface to downloading analysis-ready data cubes instead of image files from portals. 



